---
title: "All vs ORD for Channels 13 and 21"
author: "Lia Ossanna"
date: "2/2/2022"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(dplyr)
library(ggplot2)
load("ORD-vs-all.RData")
```

# Introduction
- Channels 13 and 21 have both baffles and one rock dams (ORDs). Unlike ORDs, baffles are meant to increase sinuosity by eroding the side of the channel opposite to the baffle. ORDs are meant to catch sediment and increase vegetation cover.
- Because of this distinction, I looked at the difference between including all stations in analysis for Channels 13 and 21 (baffle and ORD stations), versus ORD only. 
- I ran statistics (comparison of means) for total plant cover, soil cover, and herbaceous cover, using ANOVA with post-hoc Tukey for normally distributed comparisons, and Kruskal-Wallis with post-hoc Dunn for non-normally distributed comparisons.

# Summary
- There are no major differences except for herbaceous cover of Channel 21. The distribution was the same using all stations and ORD only, so I am unsure why the letters are so different.

# Total plant cover
```{r}
total.all1321.plot
total.ord1321.plot
```

## Stats
### All
- Channels 13 & 21 were not normally distributed.
- Channel 21 had no significant differences over time.
```{r}
total13.letters
```

### ORD only
- Channels 13 & 21 were normally distributed.
- Neither had significant differences over time.

### Comparison
- Using all stations, Channel 13 was not normally distributed and had significant changes.
- Using only ORD stations, Channel 13 was normally distributed and had no significant changes.

# Soil cover
```{r}
ground.grs1321.plot
ground.ord.grs1321.plot
```

## Stats
### All
- Channel 13 was normally distributed, while Channel 21 was not.
```{r}
hsd.soil13$groups
soil21.letters
```

### ORD only
- Channels 13 and 21 were normally distributed.
```{r}
hsd.soil.ord13$groups
hsd.soil.ord21$groups
```

### Comparison
- Using all stations, Channel 21 was not normally distributed, while using ORD stations only, Channel 21 was normally distributed.
- Channel 13 was normally distributed using all stations and ORD only stations.
- The letters were the same for both Channel 13 and 21.

# Herbaceous cover
```{r}
woody.all1321.plot
woody.ord1321.plot
```

## Stats
### All
- Channel 13 was not normally distributed; Channel 21 was normally distributed.
```{r}
herb13.letters
hsd.herb21$groups
```

### ORD only
- Channel 13 was not normally distributed; Channel 21 was normally distributed.
```{r}
herb.ord13.letters
hsd.herb.ord21$groups
```

### Comparison
- Distributions were the same for all stations and ORD only.
- Letters for both channels were slightly different. Unsure why Channel 21 ORD only are all assigned group a.
