---
title: "Analysis by Treatment3 (Treated/Control)"
author: "Lia Ossanna"
date: "2023-05-26"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(agricolae)
library(ggpubr)
load("../RData/CV_veg-2012-2021.RData")
load("../RData/ANOVA-by-Treatment3_veg-2012-2021.RData")
load("../RData/Percent-change-over-time_veg-2012-2021.RData")
```

# Introduction
- `Treatment3` has only two categories, `Treated`, and `Control`. It is a measure of the local effect of rock structures.
  - `Treated` = Channel 13 & Channel 21
  - `Control` = Channel 12 & Channel 19
- Presented here are figures and results, but not code.
- See other documents for code and analysis:
  - Temporal veg cover and diversity: `ANOVA-by-Treatment3_veg-2012-2021.html`.
  

## Averages over time
- Letters represent one-way ANOVA and Tukey HSD.
- In comparing means from linear models with `Year` as a random factor, there was no significant difference for notree cover. Control > Treated for perennial richness, and something weird happened with the Shannon diversity linear model.

```{r, echo=FALSE, fig.dim=c(11,7)}
ggarrange(notree.plot, rich.plot, shan.plot,
          ncol = 2, nrow = 2)
```

## CV
- Ran T-test for normally distributed and Wilcoxon test for non-normally distributed CV values. No significant differences between Control and Treated.
```{r, echo=FALSE, fig.dim=c(9,7)}
ggarrange(notree.plot.cv, rich.plot.cv, shan.plot.cv,
          ncol = 2, nrow = 2)
```

## Percent difference
```{r, echo=FALSE, fig.dim=c(9,7)}
ggarrange(notree.plot.pd1, rich.plot.pd1, shan.plot.pd1,
          ncol = 2, nrow = 2)
```



