---
title: "Exploratory SEMs"
author: "Lia Ossanna"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 12)
library(tidyverse)
library(lavaan)
library(semPlot)
load("../RData/Exploratory-SEM.RData")
```

# Introduction
- This document explores preliminary structural equation models (SEMs) using different response types of plant cover and diversity.
- Results are presented first, with the code example at the end.
- The left column of SEM diagrams is based on 2021 values. Right column is the difference between 2012 and 2021 (positive values indicate an increase over time).
  - The main response variable is located in the lower lefthand corner of the diagrams; this is what changes between columns.
- Variables:
  - `Width` (`Wdt`): channel width, derived from the standard deviation of LiDAR data, which measure elevation difference from 2011 to 2019; measurement is not very accurate.
  - `up.trt` (`up.`): a categorical variable, indicating if upland treatments were present (Channel 19).
  - `inch.trt` (`in.`): a categorical variable, indicating if in-channel treatments were present (Channels 13 and 21).
  - `Elev_Diff` (`E_D`): elevation difference from 2011 to 2019, as measured by LiDAR. Some stations could not be measured, so including this variable means the sample size decreases.
  - `OM_perc` (`OM_`): organic matter percent from 2021 soil samples.
  - `TN_log` (`TN_`): total nitrogen content from 2021 soil samples, log-transformed to improve normality.
- Response variables:
  - Total cover
    - `Cvr`: 2021 total cover.
    - `tt.`: difference between 2012 and 2021 total cover.
  - Herbaceous cover
    - `Hrb`: 2021 herbaceous cover.
    - `hr.`: difference between 2012 and 2021 herbaceous cover.
  - Woody cover
    - `Wdy`: 2021 woody cover.
    - `wd.`: difference between 2012 and 2021 woody cover.
  - Perennial species richness
    - `rch`: 2021 richness.
    - `rc.`: difference between 2012 and 2021 richness.
  - Perennial Shannon diversity
    - `shn`: 2021 Shannon diversity
    - `sh.`: difference between 2012 and 2021 Shannon diversity.
- The data sheet for Channel 19 Station 2 + 4 is missing for 2012, so that station is dropped from all models that use the difference between 2012 and 2021.


# Model 5-6: `Width`, `up.trt`, `inch.trt`, `Elev_Diff`, `OM_perc`, `TN_log`
- All variables used.

## Total cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.total05, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.total06, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Herbaceous cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.herb05, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.herb06, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Woody cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.wood05, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.wood06, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Richness
```{r}
par(mfrow = c(1, 2))
semPaths(fit.rich05, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.rich06, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Shannon diversity
```{r}
par(mfrow = c(1, 2))
semPaths(fit.shan05, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.shan06, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```



# Model 9-10: `up.trt`, `inch.trt`, `OM_perc`, `TN_log`
- No GIS-derived variables included.

## Total cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.total09, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.total10, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Herbaceous cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.herb09, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.herb10, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Woody cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.wood09, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.wood10, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Richness
```{r}
par(mfrow = c(1, 2))
semPaths(fit.rich09, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.rich10, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Shannon diversity
```{r}
par(mfrow = c(1, 2))
semPaths(fit.shan09, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.shan10, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```



# Model 1-2: `Width`, `up.trt`, `inch.trt`, `Elev_Diff`, `OM_perc`
## Total cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.total01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.total02, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Herbaceous cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.herb01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.herb02, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Woody cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.wood01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.wood02, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Richness
```{r}
par(mfrow = c(1, 2))
semPaths(fit.rich01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.rich02, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Shannon diversity
```{r}
par(mfrow = c(1, 2))
semPaths(fit.shan01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.shan02, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```



# Model 3-4: `Width`, `inch.trt`, `Elev_Diff`, `OM_perc`
## Total cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.total03, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.total04, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Herbaceous cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.herb03, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.herb04, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Woody cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.wood03, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.wood04, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Richness
```{r}
par(mfrow = c(1, 2))
semPaths(fit.rich03, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.rich04, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Shannon diversity
```{r}
par(mfrow = c(1, 2))
semPaths(fit.shan03, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.shan04, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```



# Model 7-8: `Width`, `inch.trt`, `Elev_Diff`, `OM_perc`, `TN_log`
## Total cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.total07, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.total08, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Herbaceous cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.herb07, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.herb08, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Woody cover
```{r}
par(mfrow = c(1, 2))
semPaths(fit.wood07, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.wood08, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Richness
```{r}
par(mfrow = c(1, 2))
semPaths(fit.rich07, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.rich08, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```

## Shannon diversity
```{r}
par(mfrow = c(1, 2))
semPaths(fit.shan07, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
semPaths(fit.shan08, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```




# Code example
```{r, eval=FALSE, echo=TRUE}
library(tidyverse)
library(lavaan)
library(semPlot)

# Load data ---------------------------------------------------------------

dat.2021 <- read.csv("data/cleaned/SEM-input.csv")
dat.2021$inch.trt <- as.factor(dat.2021$inch.trt)
dat.2021$up.trt <- as.factor(dat.2021$up.trt)
```

```{r, fig.width=7, echo=TRUE}
# Total cover -------------------------------------------------------------

# Model 1 (2021 cover)
mod.total01 <- '
  # regressions
    Cover ~ Width + up.trt + inch.trt + Elev_Diff + OM_perc
    OM_perc ~ up.trt + inch.trt
'
fit.total01 <- sem(mod.total01, data = dat.2021, fixed.x = TRUE) 
  # Warning message discussed further here: https://groups.google.com/g/lavaan/c/FzShzsXYNco?pli=1
vartable(fit.total01)
summary(fit.total01, standardized = TRUE)

semPaths(fit.total01, "std", edge.label.cex = 1.3, residuals = FALSE, sizeMan = 7)
```
